### 2025-10-01 YEAR2/MAT292_LectureNotes/main.tex
30 sept

\paragraph{General Planer Systems} Consider the system of ODEs:
$$\textbf{x}' = A\textbf{x}$$
where \( A \in \mathbb{R}^{2 \times 2} \).

>>>>

\paragraph{General Planer Systems} Consider the system of ODEs:
$$\textbf{x}' = A\textbf{x}$$
where \( A \in \mathbb{R}^{2 \times 2} \).
We define the fundamental matrix \( \Phi(t) \) as:
\begin{equation}
\Phi(t) = \begin{bmatrix}
x_1(t) & x_2(t) \\ 
y_1(t) & y_2(t)
\end{bmatrix}
\end{equation}
where \( (x_1(t), y_1(t)) \) and \( (x_2(t), y_2(t)) \) are two linearly independent solutions of the system. The Wronskian of the system is given by:
\begin{equation}
W(t) = \det(\Phi(t)) = x_1(t)y_2(t) - x_2(t)y_1(t)
\end{equation}
The Wronskian satisfies the differential equation:
\begin{equation}
W'(t) = \text{tr}(A) W(t)
\end{equation}
with the solution:
\begin{equation}
W(t) = W(0) e^{\text{tr}(A)t}
\end{equation}
where \( W(0) \) is the Wronskian at \( t = 0 \).

\subsection{Stability of Nonlinear Systems in Local Linear Approximation}
\begin{definition}[Change of Variables]
    The change of variable in a higher dimensions, the change of variables related by the Jacobian matrix. WLOG, consider change of variables $F: \mathbb{R}^2 \to \mathbb{R}^2$ and system $x' = Ax$. Let:
    $$
    F(x,y) = \begin{bmatrix}f_1(x,y) \\ f_2(x,y)
    \end{bmatrix}
    $$

    Then, the Jacobian matrix \( J_F \) is given by:
    \begin{equation}
    J_F = \begin{bmatrix}
    \frac{\partial f_1}{\partial x} & \frac{\partial f_1}{\partial y} \\
    \frac{\partial f_2}{\partial x} & \frac{\partial f_2}{\partial y}
    \end{bmatrix}
    \end{equation}
    The change of variables transforms the system \( x' = Ax \) into a new system \( u' = Bu \) where:
    \begin{equation}
    B = J_F A J_F^{-1}
    \end{equation}
    provided that \( J_F \) is invertible.
\end{definition}

\begin{shaded}
\subsubsection*{A reference for simple linear stability analysis.}

\paragraph{Linear Stablity Analysis} Consider the simple linear one-dimensional ODE:
$$y' = f(y)$$
where \( f(y) \) is a differentiable function. Let \( y^* \) be a fixed point, i.e., \( f(y^*) = 0 \). We can perform a linear stability analysis by considering a small perturbation \( \eta(t) = y(t) - y^* \). Then, we have:
\begin{align*}
\eta' &= y' - (y^*)' \\
&= f(y) - 0 \\
&= f(y^* + \eta) \\
&\approx f(y^*) + f'(y^*)\eta \\
&= f'(y^*)\eta = A\eta
\end{align*}
So, we have the linearized system:
$$\eta' = A\eta$$
where \( A = f'(y^*) \). The stability of the fixed point \( y^* \) is determined by the sign of \( A \).
\end{shaded}

\begin{definition}[Linear Stability Analysis in Higher Dimensions]
Consider the two-dimensional system:
$$\begin{cases}
x' = f(x, y) \\
y' = g(x, y)
\end{cases}$$
where \( f(x, y) \) and \( g(x, y) \) are differentiable functions. Let \( (x^*, y^*) \) be a fixed point. We let perturbations \( u(t) = x(t) - x^* \) and \( v(t) = y(t) - y^* \). Then, we have:
$$
\begin{cases}
    u' &= x' = \cancel{f(x,y)} + \frac{\partial f}{\partial x}(x^*, y^*)u + \frac{\partial f}{\partial y}(x^*, y^*)v + \cancel{O(u^2, v^2, uv)} \\
    v' &= y' = \cancel{g(x,y)} + \frac{\partial g}{\partial x}(x^*, y^*)u + \frac{\partial g}{\partial y}(x^*, y^*)v + \cancel{O(u^2, v^2, uv)}
\end{cases}
$$
So we can write the linearized system as:
$$\begin{bmatrix}
u' \\ v'
\end{bmatrix} = J \begin{bmatrix}
u \\ v
\end{bmatrix}$$
where \( J \) is the Jacobian matrix evaluated at the fixed point \( (x^*, y^*) \). This describes the local behavior of the system near the fixed point. Then the stability and behavior is determined by the determinant and trace of the Jacobian matrix \( J \), and hence the eigenvalues and eigenvectors of \( J \).
\end{definition}

\begin{definition}
    Consider the nonlinear pendulum equation:
    $$\theta'' + \frac{g}{L}\sin(\theta) = 0$$
    where \( \theta \) is the angle of the pendulum, \( g \) is the acceleration due to gravity, and \( L \) is the length of the pendulum. We can rewrite this as a system of first-order ODEs by defining:
    $$\begin{cases}
    x_1 = \theta \\
    x_2 = \theta'
    \end{cases}$$
    Then, we have:
    $$\begin{cases}
    x_1' = x_2 \\
    x_2' = -\frac{g}{L}\sin(x_1)    
    \end{cases}$$
    Now, consider the Jacaobian matrix:
    $$J = \begin{bmatrix}
    0 & 1 \\
    -\frac{g}{L}\cos(x_1) & 0
    \end{bmatrix}$$

    Consider the fixed point $(n \pi, 0)$ for \( n \in \mathbb{Z} \). We have:
    $$J = \begin{bmatrix}
    0 & 1 \\
    -\frac{g}{L}\cos(n \pi) & 0
    \end{bmatrix} = \begin{bmatrix}
    0 & 1 \\
    -\frac{g}{L}(-1)^n & 0
    \end{bmatrix}$$
    So the derminant is:
    $$\det(J) = -\frac{g}{L}(-1)^n$$
    and the trace is:
    $$\text{tr}(J) = 0$$
    When $n$ is even, we have a positive determinant, so the fixed point is a center. When $n$ is odd, we have a negative determinant, so the fixed point is a saddle node.
\end{definition}

### 2025-10-01 YEAR2/PHY293_LectureNotes/main.tex
30 sept

\paragraph{Wave boundry Conditions} When a wave encounters a boundary between two different media, part of the wave is reflected back into the original medium, and part is transmitted into the new medium. This is constrined by the following boundary conditions:
\begin{itemize}
    \item The displacement of the wave must be continuous across the boundary:
    \begin{equation}
        y_1(x_b, t) = y_2(x_b, t) = y_3(x_b, t)
    \end{equation}
    \item The derivative of the displacement with respect to position must also be continuous across the boundary:
    \begin{equation}
        \frac{\partial y_1}{\partial x}\bigg|_{x_b} = \frac{\partial y_2}{\partial x}\bigg|_{x_b} = \frac{\partial y_3}{\partial x}\bigg|_{x_b}
    \end{equation}
\end{itemize}
and then from these conditions, we can derive the amplitudes of the reflected and transmitted waves:
\begin{equation}
    A_i = A_r + A_t
\end{equation}
and 
\begin{equation}
    A_i Z_1 = A_r Z_1 + A_t Z_2
\end{equation}
From these two equations, we can solve for the reflection and transmission coefficients:
\begin{equation}
    R = \frac{A_r}{A_i} = \frac{Z_2 - Z_1}{Z_2 + Z_1}
\end{equation}
and
\begin{equation}
    T = \frac{A_t}{A_i} = \frac{2Z_2}{Z_2 + Z_1}
\end{equation}
\chapter{Modern Physics}
\end{document}

>>>>

\paragraph{Wave boundry Conditions} When a wave encounters a boundary between two different media, part of the wave is reflected back into the original medium, and part is transmitted into the new medium. This is constrined by the following boundary conditions:
\begin{itemize}
    \item The displacement of the wave must be continuous across the boundary:
    \begin{equation}
        y_1(x_b, t) = y_2(x_b, t) = y_3(x_b, t)
    \end{equation}
    \item The derivative of the displacement with respect to position must also be continuous across the boundary:
    \begin{equation}
        \frac{\partial y_1}{\partial x}\bigg|_{x_b} = \frac{\partial y_2}{\partial x}\bigg|_{x_b} = \frac{\partial y_3}{\partial x}\bigg|_{x_b}
    \end{equation}
\end{itemize}
and then from these conditions, we can derive the amplitudes of the reflected and transmitted waves:
\begin{equation}
    A_i = A_r + A_t
\end{equation}
and 
\begin{equation}
    A_i Z_1 = A_r Z_1 + A_t Z_2
\end{equation}
From these two equations, we can solve for the reflection and transmission coefficients:
\begin{equation}
    R = \frac{A_r}{A_i} = \frac{Z_2 - Z_1}{Z_2 + Z_1}
\end{equation}
and
\begin{equation}
    T = \frac{A_t}{A_i} = \frac{2Z_2}{Z_2 + Z_1}
\end{equation}

\paragraph{Special Cases} There are two special cases:
\begin{itemize}
    \item \textbf{Fixed End}: If the wave encounters a boundary where the medium is fixed (e.g., a string tied to a wall), the reflected wave undergoes a phase change of \( \pi \) (inversion). The reflection coefficient is \( R = -1 \) and the transmission coefficient is \( T = 0 \).
    \item \textbf{Free End}: If the wave encounters a boundary where the medium is free to move (e.g., a string attached to a ring that can slide along a rod), the reflected wave does not undergo a phase change. The reflection coefficient is \( R = 1 \) and the transmission coefficient is \( T = 2 \).
\end{itemize}

\begin{example}[Transmission of 64\% Energy]
Suppose a wave transmits \(64\%\) of its incident energy across a boundary. Then:
\begin{equation}
    T_e = 0.64, 
    \quad R_e = 1 - T_e = 0.36
\end{equation}
so that the amplitude reflection coefficient is
\begin{equation}
    R = +\sqrt{R_e} = 0.6 
\end{equation}
\textbf{Important} the positive sign indicates the reflected wave is \emph{in phase} with the incident wave.

From the boundary condition in Eq.~\eqref{eq:amp-continuity},
\begin{equation}
    T = 1 + R = 1 + 0.6 = 1.6
\end{equation}
Thus, the amplitude transmission coefficient is \(T = 1.6\), even though only \(64\%\) of the energy is transmitted. This apparent paradox arises because the transmitted energy depends on both the amplitude squared and the ratio of impedances \(\tfrac{Z_2}{Z_1}\).
\end{example}
\subsection{Superposition and Standing Waves}
\begin{definition}[Modulation of two waves]
    Consider two monochromatic waves:
    $$
    \begin{cases}
        y_1(x,t) &= A \cos(k_1 x - \omega_1 t) \\
        y_2(x,t) &= A \cos(k_2 x - \omega_2 t)
    \end{cases}
    $$
    Then, the resultant wave $y_1 + y_2$ can be expressed as:
    \begin{equation}
        y(x,t) = 2A \cos\left(\frac{(k_1 - k_2)x - (\omega_1 - \omega_2)t}{2}\right) \cos\left(\frac{(k_1 + k_2)x - (\omega_1 + \omega_2)t}{2}\right)
    \end{equation}
\end{definition}

\begin{definition}[Coherence]
    Two waves are said to be coherent if they are monochromatic and maintain a constant phase relationship over time. This means that the difference in phase between the two waves does not change as they propagate:
    \begin{equation}
        \Delta \phi = \phi_1 - \phi_2 = \text{constant}
    \end{equation}
    Coherent waves can produce stable interference patterns, such as constructive and destructive interference.
\end{definition}

\begin{definition}[Phase and Phase Difference]
    Phase of the wave is an argument $kx \pm \omega t + \phi_0$ of the wave function. The phase difference between two waves at a given point in space and time is given by:
    \begin{equation}
        \Delta \phi = (k_1 x - \omega_1 t + \phi_1) - (k_2 x - \omega_2 t + \phi_2)
    \end{equation}
    where \( k_1, k_2 \) are the wave numbers, \( \omega_1, \omega_2 \) are the angular frequencies, and \( \phi_1, \phi_2 \) are the initial phases of the two waves.

    If we consider two waves of the same frequency and wave number, the phase difference simplifies to:
    \begin{equation}
        \Delta \phi = \Delta\phi + k\Delta x + \omega \Delta t
    \end{equation}

\end{definition}
\begin{definition}[Waves in Two and Three Dimensions]
    In two or three dimensions, waves can propagate in various directions. The general form of a wave in two dimensions is:
    \begin{equation}
        z(x,y,t) = A \cos(k_x x + k_y y - \omega t + \phi_0)
    \end{equation}
    where \( k_x \) and \( k_y \) are the wave numbers in the \( x \) and \( y \) directions, respectively, and \( \phi_0 \) is the initial phase. We also define $v$ and $k$ as:
    \begin{equation}
        v = \frac{\omega}{k}, \quad k = \sqrt{k_x^2 + k_y^2}
    \end{equation}
\end{definition}

\begin{definition}[Spherical Waves]
    A spherical wave is a wave that propagates outward in all directions from a point source. The general form of a spherical wave is:
    \begin{equation}
        y(r,t) = \frac{A}{r} \cos(kr - \omega t + \phi_0)
    \end{equation}
    where \( r = \sqrt{x^2 + y^2 + z^2} \) is the radial distance from the source, \( A \) is the amplitude at a reference distance, \( k \) is the wave number, \( \omega \) is the angular frequency, and \( \phi_0 \) is the initial phase. The amplitude decreases with distance as \( 1/r \) due to the spreading of the wavefront.

\begin{definition}[Interference]
    Consider two coherent in phase sources $S_1$ and $S_2$ separated by a distance \( d \). The path difference \( \Delta r \) between the two waves at a point \( P \) is given by:
    \begin{equation}
        \Delta r = r_2 - r_1
    \end{equation}
    where \( r_1 \) and \( r_2 \) are the distances from the sources \( S_1 \) and \( S_2 \) to the point \( P \), respectively.

    We have the following two cases for constructive and destructive interference:
    \begin{itemize}
        \item \textbf{Constructive Interference}: Occurs when the path difference is an integer multiple of the wavelength:
        \begin{equation}
            \Delta r = m \lambda, \quad m = 0, 1, 2, \ldots
        \end{equation}
        \item \textbf{Destructive Interference}: Occurs when the path difference is an odd multiple of half the wavelength:
        \begin{equation}
            \Delta r = \left(m + \frac{1}{2}\right) \lambda, \quad m = 0, 1, 2, \ldots
        \end{equation}
    \end{itemize}
\end{definition}

\chapter{Modern Physics}
\end{document}
